% anomaly-search.tex — paper/anomaly-search-full.md の LaTeX 化 (v18.x 出版作業)
% 対象誌: SciPost Physics / CPC / PRD (revtex4-2, preprint)
% 数値の一次ソースは results/ と certificates/ — 本文数値は .md 原稿から転記
% TODO(著者確認): 著者名・所属は仮置き (リポジトリ所有者)。投稿前に確定すること。
\documentclass[aps,prd,preprint,nofootinbib]{revtex4-2}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{amsmath,amssymb}
\usepackage{booktabs}
\usepackage{hyperref}

\begin{document}

\title{The Standard Model as the Unique Minimal Anomaly-Free Chiral Spectrum:\\
An Exhaustive, Independently Cross-Checked, and Machine-Verified Search}

\author{Yasuo Higano}
\affiliation{Independent researcher; Quantum Relational Network program}

\date{\today}

\begin{abstract}
We present an exhaustive search for anomaly-free chiral fermion spectra of
$SU(3)\times SU(2)\times U(1)$ gauge theories within explicitly bounded domains
(representations up to the color octet and weak triplet, hypercharges $|Y|\le 3$,
up to 8 multiplets), subject to the five perturbative anomaly-cancellation
conditions, the Witten $SU(2)$ anomaly, chirality, and charge under all factors.
Within every domain scanned, the minimal solution is the 15-component Standard
Model generation; it is unique at 15 components and isolated above them. A
negative-control map identifies which conditions carry which conclusion:
minimality is enforced by chirality, all-factor charge, the Witten anomaly, and
the cubic $SU(3)$ anomaly, while uniqueness is enforced by the cubic $U(1)$
anomaly --- the three linear anomaly conditions are redundant within the domain,
itself a finding. The search is implemented three times independently
(depth-first, linear-elimination grid, meet-in-the-middle), certified by SHA-256
canonical-form certificates, and the two core domains are verified as theorems
in Lean~4 via \texttt{native\_decide} with a poisoned-fuel design that makes
completeness of the enumeration part of the theorem statement. Extending the
gauge group: the only rank-2 chiral extension in the window is $B-L$ (requiring
$\nu_R$, unique as a charge plane), no rank-3 chiral extension exists (0
solutions against 355 in the vectorlike control), and among exceptional groups
only $E_6$ admits chiral matter --- its $\mathbf{27}$ reducing exactly to one SM
generation plus vectorlike remainder. Consistency alone thus counts the forces
and fixes the matter, within windows whose boundaries we state precisely. All
results are reproducible from a dependency-free Rust repository with a
machine-validated claim ledger.
\end{abstract}

\maketitle

\section{Introduction}

Why this matter content? The Standard Model (SM) generation --- fifteen Weyl
components in five multiplets with hypercharges
$(1,1)_{-6}\oplus(1,2)_{3}\oplus(3,1)_{-2}\oplus(3,1)_{4}\oplus(\bar 3,2)_{-1}$
(in units of $Y/6$) --- satisfies a tightly coupled system of
anomaly-cancellation conditions. That the SM is anomaly-free is textbook
material; that hypercharge is \emph{forced} by anomaly cancellation under stated
assumptions has a long literature~\cite{GengMarshak,BabuMohapatra,Foot}. This
paper asks and answers a computational version of the strongest form of the
question:

\begin{quote}
Within an explicitly bounded space of chiral gauge theories, is the SM
generation the \textbf{minimal} consistent spectrum? Is it \textbf{unique} at
its size? Is it \textbf{isolated} above it? And which consistency conditions
carry each of these properties?
\end{quote}

Our answers: yes, yes, yes --- with the entire component-count spectrum
$\{15{:}\,1,\ 16{:}\,8,\ 17{:}\,1,\ 18{:}\,18,\ 22{:}\,2,\ 24{:}\,459\}$
enumerated in the widest domain, and with a \emph{control map} attributing
minimality and uniqueness to specific conditions. The method is exhaustive
integer enumeration: no sampling, no heuristics, exact arithmetic throughout,
with completeness itself made checkable (\S\ref{sec:methods}).

Three methodological features distinguish this from a table of known facts.
First, \emph{independent triple implementation}: the same domains are
enumerated by three algorithms sharing no code path, and their canonical-form
solution sets agree byte-for-byte (SHA-256 certificates in the repository).
Second, \emph{negative controls}: each consistency condition is switched off in
turn, and the search is re-run to see what breaks --- the instrument is
demonstrated to be able to fail. Third, \emph{machine verification}: the two
core domains are re-proved inside Lean~4, with a fuel-poisoning construction
that converts ``the enumeration did not silently skip a branch'' into part of
the theorem.

\section{Setup}

A candidate spectrum is a multiset of at most $k$ multiplets $(R_c,R_w)_y$ with
$R_c\in\{1,3,\bar 3,6,\bar 6,8\}$, $R_w\in\{1,2,3\}$, and integer $y=6Y$ in a
stated window. The consistency conditions are:
\begin{itemize}
\item A1--A3 (linear): $[SU(3)]^2 U(1)$, $[SU(2)]^2 U(1)$, $\mathrm{grav}^2 U(1)$;
\item A4: $[U(1)]^3$;\quad A5: $[SU(3)]^3$;
\item W: Witten's $SU(2)$ global anomaly~\cite{Witten} (even number of weak
      doublets, counted with color dimension);
\item C: chirality (no multiplet together with its conjugate; no completely
      real content);
\item F: charged under all three factors (some color, some weak, some
      hypercharge non-trivially present).
\end{itemize}

Domains (machine-readable in \texttt{certificates/v62\_domains.json}): the core
domain $D_1$ (small representations, $|Y|\le 3/2$, $\le 5$ multiplets, $\le 15$
components), the extension domain $D_2$ ($\le 6$ multiplets, $\le 24$
components), and robustness axes: large representations ($6$, $\bar 6$, $8$;
weak triplets), $|Y|\le 2$ and $\le 3$, up to 8 multiplets, and an added
right-handed neutrino. The physical reading of the window is an EFT assumption
--- heavy vectorlike matter decouples (verified separately at $M^{-1.91}$ in
this program) --- so the theorems are of the form ``within this window,''
stated without apology and without extrapolation.

\section{Results: minimality, uniqueness, isolation}

\textbf{Minimality and uniqueness.} In every domain scanned, the minimal chiral
solution has 15 components and is exactly the SM generation, unique up to the
canonical equivalences (permutation, global conjugation, $U(1)$ sign, and
charge rescaling). This holds with large representations admitted, with $|Y|$
up to 3, with up to 8 multiplets, and with $\nu_R$ present (where the
16-component SM$+\nu_R$ appears as expected at the next size).

\textbf{The spectrum.} The widest domain yields component counts
$\{15{:}\,1,\ 16{:}\,8,\ 17{:}\,1,\ 18{:}\,18,\ 22{:}\,2,\ 24{:}\,459\}$: one
solution at 15; eight 16-component neighbors, each excluded empirically by
fractionally charged hadrons or similar observational inputs (this exclusion is
observational, not group-theoretic --- we mark it as such); a single
17-component solution containing a weak triplet; and nothing between 18 and 22.
The SM is not merely minimal; it sits in a \emph{gap}.

\textbf{The control map.} Re-running with individual conditions removed
(Table~\ref{tab:controls}): the linear anomaly conditions, so prominent in
textbook derivations, are \emph{implied} by the remaining conditions within the
domain; minimality is carried by chirality $+$ all-factor charge $+$ Witten $+$
cubic $SU(3)$, and uniqueness by the cubic $U(1)$. The baseline solution set is
contained in every control's solution set (monotonicity check), so the controls
genuinely relax rather than shift the problem.

\begin{table}[t]
\caption{\label{tab:controls}The negative-control map: each consistency
condition is removed in turn and the search re-run.}
\begin{ruledtabular}
\begin{tabular}{lccl}
condition removed & minimal size & \#\,at 15 & reading \\
\colrule
chirality           & 10 & 59 & minimality \emph{and} uniqueness collapse \\
all-factor charge   & 5  & 1  & minimality collapses (pure-$U(1)$ towers) \\
Witten $SU(2)$      & 12 & 1  & minimality collapses \\
$[SU(3)]^3$         & 12 & 6  & both degrade \\
$[SU(3)]^2U(1)$     & 15 & 1  & \textbf{redundant in-domain} \\
$[SU(2)]^2U(1)$     & 15 & 1  & \textbf{redundant in-domain} \\
$\mathrm{grav}^2U(1)$ & 15 & 1 & \textbf{redundant in-domain} \\
$[U(1)]^3$          & 15 & 9  & uniqueness collapses \\
\end{tabular}
\end{ruledtabular}
\end{table}

\section{Methods: three implementations, certificates, machine verification}
\label{sec:methods}

\textbf{Triple implementation.} (i) Depth-first with pruning over multiplet
stacks; (ii) linear-elimination over a hypercharge grid; (iii)
meet-in-the-middle over half-assignments with exact integer keys (the anomaly
conditions split additively). All arithmetic is exact
(\texttt{i64}/\texttt{i128}); canonical forms quotient permutations,
conjugation, $U(1)$ flips, and gcd rescaling. The three solution sets agree
exactly in every domain; their canonicalized listings are certified by SHA-256
(\texttt{certificates/}), and CI recomputes and compares the hashes.

\textbf{Machine verification in Lean 4.} The core domains are re-proved as
theorems of the form $\texttt{scanResult}=4$ (the number of canonical-orbit
representatives, SM's orbit): the enumeration is a structurally recursive walk
whose fuel is \emph{poisoned} --- running out of fuel contributes $+10^9$ and
an out-of-orbit hit contributes $+10^6$ to the result --- so the equality
theorem simultaneously asserts correctness \emph{and completeness} of the walk.
\texttt{native\_decide} evaluates the closed decision procedure (991\,s for the
small-representation domain; 9{,}734\,s for the large-representation domain,
$1.7\times 10^{10}$ recursive calls). The trust base is Lean's kernel plus the
native evaluator, which we state explicitly; the Rust and Lean enumerations are
independent codings of the same domain definition.

\textbf{Runtime prediction.} Lean \texttt{native\_decide} wall time is
proportional to the total number of recursive calls ($\sim 0.6\,\mu$s/call),
measured by instrumenting the identical recursion in Rust first. We recommend
this practice: enumeration-completeness proofs are affordable exactly when
their cost can be predicted before writing them.

\section{Climbing the gauge group: the U(1) staircase and the exceptional ceiling}

\textbf{Second $U(1)$.} With the SM matter fixed and a second hypercharge $Y'$
scanned over $|6Y'|\le 12$: without $\nu_R$ every solution is proportional to
$Y$; with $\nu_R$ the solutions span exactly the $\{Y,\,B-L\}$ plane. A
structure-free rank-2 scan (both charges free, $\le 5$--$6$ multiplets) finds
\emph{no} chiral rank-2 solution below 16 components, and at 16 exactly one
charge \emph{plane} --- $\{Y,\,B-L\}$ on SM$+\nu_R$ --- after
Pl\"ucker-coordinate classification of planes ($GL(2,\mathbb{Z})$ lattice
classification overcounts bases of the same plane; we document this
initially-wrong equivalence and its fix). \textbf{The minimal way to add a
second force is $B-L$, and it forces the right-handed
neutrino}~\cite{BabuMohapatra,Foot,Allanach,Costa}.

\textbf{Third $U(1)$.} A rank-3 scan (19 exact conditions: 9 linear, 10 cubic
monomials) over the stated window finds \textbf{zero} chiral solutions; the
vectorlike control finds 355, so the instrument is not blind. The staircase
ends at two steps: $Y$, then $B-L$, then nothing.

\textbf{Exceptional groups.} $G_2$, $F_4$, $E_7$, $E_8$ admit only
self-conjugate representations; we verify $W=-W$ as exact integer
weight-multiset identities for $\mathbf{7}$, $\mathbf{14}$, $\mathbf{26}$,
$\mathbf{52}$, $\mathbf{56}$, $\mathbf{133}$, $\mathbf{248}$ (with Weyl-closure
checks of each construction, and $SU(3)$'s complex $\mathbf{3}$ as the negative
control), so their chiral cores are empty. Among exceptional groups only $E_6$
has complex representations, and the chiral core of its $\mathbf{27}$ --- after
stripping conjugate pairs and singlets exactly --- is one SM generation. The
staircase and the exceptional ceiling point at the same spectrum.

\section{Limitations}

The windows are finite and stated: large-$|Y|$ rank-3 territory, larger
multiplet counts, and huge representations remain open; the 16-component
neighbors are excluded by observational inputs, not consistency; the
exceptional-group statement verifies representative low-dimensional
representations and cites the standard self-conjugacy theorems for the
universal part; the v4.3-domain Lean formalization at $\sim 10^{11}$ calls
awaits an array-based enumeration design. None of these limitations, if lifted,
can \emph{remove} solutions --- every extension so far has only confirmed the
gap around the SM.

\section{Reproducibility}

Dependency-free Rust (std only), fixed seeds, exact integer arithmetic in all
searches, [PASS]/[FAIL] verification embedded in every program (314 PASS / 0
FAIL across the repository at time of writing), SHA-256 certificates for all
solution sets, machine-validated claim ledger (94 claims with grades, evidence
files, and limitations), and CI that rebuilds, re-runs the audit suite,
re-validates certificates, and re-checks both Lean theorems.

\begin{thebibliography}{9}
\bibitem{Witten} E.~Witten, ``An $SU(2)$ anomaly,'' Phys.\ Lett.\ B \textbf{117} (1982) 324.
\bibitem{GengMarshak} C.~Q.~Geng and R.~E.~Marshak, ``Uniqueness of quark and lepton representations in the standard model from the anomalies viewpoint,'' Phys.\ Rev.\ D \textbf{39} (1989) 693; J.~A.~Minahan, P.~Ramond and R.~C.~Warner, ``A comment on anomaly cancellation in the standard model,'' Phys.\ Rev.\ D \textbf{41} (1990) 715.
\bibitem{BabuMohapatra} K.~S.~Babu and R.~N.~Mohapatra, ``Is there a connection between quantization of electric charge and a Majorana neutrino?,'' Phys.\ Rev.\ Lett.\ \textbf{63} (1989) 938.
\bibitem{Foot} R.~Foot, G.~C.~Joshi, H.~Lew and R.~R.~Volkas, ``Charge quantization in the standard model and some of its extensions,'' Mod.\ Phys.\ Lett.\ A \textbf{5} (1990) 2721.
\bibitem{Allanach} B.~C.~Allanach, B.~Gripaios and J.~Tooby-Smith, ``Anomaly cancellation with an extra gauge boson,'' Phys.\ Rev.\ Lett.\ \textbf{125} (2020) 161601.
\bibitem{Costa} D.~B.~Costa, B.~A.~Dobrescu and P.~J.~Fox, ``General solution to the U(1) anomaly equations,'' Phys.\ Rev.\ Lett.\ \textbf{123} (2019) 151601, arXiv:1905.13729.
\bibitem{Lean} The Lean~4 theorem prover; the trust base of \texttt{native\_decide} is stated explicitly in \S\ref{sec:methods}.
\bibitem{Repo} This program's repository: Quantum Relational Network (\texttt{claims.yml}, \texttt{results/}, \texttt{certificates/}, \texttt{proofs/}) --- primary source for all numbers and certificates.
\end{thebibliography}

\end{document}
